---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "A Bayesian nested patch occupancy model to estimate steelhead movement and abundance"
```

# Manuscript for Upper Columbia steelhead patch-occupancy model

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/KevinSee/ManuscptUCsthdDABOM/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Waterhouse, L., J. White, K. See, A. Murdoch, B. Semmens, (2020). _`r Title`_. Ecological Applications <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Waterhouse, L., J. White, K. See, A. Murdoch, B. Semmens, (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>


### How to cite

Please cite this compendium as:

> K. See, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: <https://github.com/KevinSee/ManuscptUCsthdDABOM/archive/master.zip>

Or you can install this compendium as an R package, ManuscptUCsthdDABOM, from GitHub with:


```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("KevinSee/ManuscptUCsthdDABOM")
```


### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

## Prerequisites

The results cited in the manuscript were derived using the **D**am **A**dult **B**ranch **O**ccupancy **M**odel (DABOM). `DABOM` is an R package for estimating abundance of anadromous fishes using fish tagged at a primary sampling facility (e.g., dam) and later detected in upstream tributaries. It incorporates the branching structure of a stream network, and simultaneously estiamtes imprefect detection probabilities at all detection locations and movement probabilities past detection locations. The movement probabilities, when multiplied correctly and combined with an estimate of total abundance at the tagging site, can be used to estimate abundance at a variety of spatial scales. 

To install `DABOM` you can use Hadley Wickham's `devtools` package. To install and load the `devtools` package use:

```
install.packages("devtools")
library(devtools)
```
NOTE: To use `devtools`, you may also have to download and install Rtools (although it may already be installed). The latest version on Rtools can be found at
https://cran.r-project.org/bin/windows/Rtools/

Once `devtools` is successfully installed, use the following to install DABOM from GitHub:
```
devtools::install_github("KevinSee/DABOM@v0.1.0")
```

DABOM requires the JAGS software (**J**ust **A**nother **G**ibbs **S**ampler). This can be downloaded here:

https://sourceforge.net/projects/mcmc-jags/files/

Please download version >= 4.0.0

## Reproducing Results

To reproduce the results from our manuscript, clone this repository and navigate to "`analysis/R_scripts`" and then run the following R scripts:

* `01_process_capture_histories.R`
* `02_run_DABOM.R`
* `03_summarise_results`

The results from comparing multiple model runs using some upstream observation sites with runs without those sites can be seen by knitting together the Rmarkdown document, `04_upstrm_compare.Rmd`.
